---
title: "Predicting Stroke"
output:
  html_document:
    toc: yes
    number_sections: yes
  word_document:
    toc: yes
---

# Introduction

## Overview

According to the World Health Organization (WHO) stroke is the 2nd leading cause of death globally, responsible for approximately 11% of total deaths. Six and a half-million people die from stroke annually(3.3 million from ischaemic stroke annually, 2.9 million from intracerebral haemorrhages and 0.3 million from subarachnoid haemorrhages). One in four people over age 25 will have a stroke in their lifetime. 62% of all incident strokes are ischaemic strokes while 28% are intracerebral haemorrhages and 10% are subarachnoid haemorrhages.


## About the Data

This data set is used to predict whether a patient is likely to get stroke based on some input parameters like gender, age, hypertension, heart diseases, residence, bmi and smoking status. Each row in the data provides relevant information about the patient.

The data dictionary is as follows:

- id: unique identifier
- gender: "Male", "Female" or "Other"
- age: age of the patient
- hypertension: 0 if the patient doesn't have hypertension, 1 if the patient has hypertension
- heart_disease: 0 if the patient doesn't have any heart diseases, 1 if the patient has a heart disease
- ever_married: "No" or "Yes"
- work_type: "children", "Govt_jov", "Never_worked", "Private" or "Self-employed"
- Residence_type: "Rural" or "Urban"
- avg_glucose_level: average glucose level in blood
- bmi: body mass index
- smoking_status: "formerly smoked", "never smoked", "smokes" or "Unknown"*
- stroke: 1 if the patient had a stroke or 0 if not

*Note: "Unknown" in smoking_status means that the information is unavailable for this patient


# Data Exploration

```{r setup, include=FALSE}
## Set Chunk requirements
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

```

## loading Relevant packages

```{r Import relevant packages}
#Import relevant packages

library(tidyverse)
library(janitor)
library(readr)
library(plotly)
library(knitr)


```


## loading Data Set


```{r importing data}
stroke <- read_csv('https://raw.githubusercontent.com/reinpmomz/Data_sets/main/Data/healthcare-dataset-stroke-data.csv', na=c("N/A"))%>%
  clean_names()

```

```{r}
rmarkdown::paged_table(stroke)
```

```{r}
dim(stroke)
```

* Our data has 5110 observations and 12 variables

```{r checking the data structure}
##checking the data structure

str(stroke)

```

From the output on the data structure, all of the data has been read as numeric values('double' value or a decimal type with at least two decimal places) and character values but some should be converted to factors since they are categorical.

## Converting into factors

```{r converting categorical to factors}

stroke_final <- stroke%>%
  filter(gender!= "Other")%>%
  mutate(gender = factor(gender, levels = c("Female", "Male")))%>%
  mutate(hypertension= factor(hypertension, levels = c(0,1), 
                                      labels = c("No", "Yes")))%>%
  mutate(heart_disease= factor(heart_disease, levels = c(0,1), 
                                      labels = c("No", "Yes")))%>%
  mutate(ever_married = factor(ever_married, levels = c("No", "Yes")))%>%
  mutate(across(c(work_type, residence_type, smoking_status), as.factor))%>%
  mutate(stroke= factor(stroke, levels = c(0,1), 
                                      labels = c("No", "Yes")))%>%
  labelled::set_variable_labels(
    id = "unique identifier",
  gender = "Sex of the patient",
  age = "Age of the patient",
  hypertension = "Patient has hypertension",
  heart_disease = "Patient has heart disease",
  ever_married = "Patient is married",
  work_type = "Work type",
  residence_type = "Residence type",
  avg_glucose_level = "Average glucose level in blood",
  bmi = "Body mass index (in kg/m2)",
  smoking_status = "Smoking status",
  stroke = "Patient had stroke"
  )

```



## checking missing values

```{r missing values}

sum(is.na(stroke_final))

#which(is.na(stroke_final))

#which(!complete.cases(stroke_final))


sapply(stroke_final,function(x) sum(is.na(x)))

```
There were 201 missing values in bmi variable in our data set.

# Exploratory data analysis


## Univariate analysis

This is analysis of one variable to enable us understand the distribution of values for a single variable.

### Normality of continous variables


```{r}
shapiro.test(stroke_final$bmi)
```

The shapiro.test has a restriction in R that it can be applied only up to a sample of size 5000 and the least sample size must be 3. Therefore, we have an alternative hypothesis test called Anderson Darling normality test. To perform this test, we need load nortest package and use the ad.test function

```{r}
nortest::ad.test(stroke_final$age)

```

```{r}
nortest::ad.test(stroke_final$avg_glucose_level)
```


If the p-value < 0.05, it implies that the distribution of the data are significantly different from normal distribution. In other words, we cannot assume the normality.


### Descriptives Frequency table

```{r Import relevant packages2}
library(gtsummary)
library(flextable)

set_gtsummary_theme(list(
  "tbl_summary-fn:percent_fun" = function(x) style_percent(x, digits = 1),
  "tbl_summary-str:categorical_stat" = "{n} ({p}%)"
))
# Setting `Compact` theme
theme_gtsummary_compact()

```


```{r descriptives}
# make dataset with variables to summarize

      
tbl_summary(stroke_final%>%
              dplyr::select(-id),
                      type = list(
                        all_dichotomous() ~ "categorical",
                         all_continuous() ~ "continuous2")
                      , statistic = all_continuous() ~ c(
                                     "{mean} ({sd})", 
                                     "{median} ({p25}, {p75})", 
                                     "{min}, {max}")
                      , digits = all_continuous() ~ 2
                      , missing = "always" # don't list missing data separately
                      ,missing_text = "Missing"
                      ) %>% 
  modify_header(label = "**Descriptives**") %>% # update the column header
  bold_labels() %>%
  italicize_levels()%>%
  add_n() # add column with total number of non-missing observations


```

### Visualization



```{r bar graphs1, out.width="50%",fig.show="hold"}
library(ggpubr)


 ggplot(stroke_final, aes(x=stroke))+
  geom_bar(aes(fill = stroke), show.legend = FALSE)+
  labs(x="",y="", title = "Patient had stroke")+
  geom_text(aes(label = paste0(..count.., " (", scales::percent(after_stat(prop), accuracy = .1), ")"), group=1),
            stat = "count", vjust = 1.1, colour = "black")+
  #guides(fill = FALSE)+
  theme_pubclean()+
  theme(axis.title = element_text(face="bold",color="black",size=13),
        #legend.position = "none",
        axis.text.y = element_text(color="black",size=12),
        axis.text.x = element_text(color="black",size=11),
        plot.title = element_text(hjust = 0.5, face="bold",color="black",size=13),
        panel.grid.major.x = element_blank())


 ggplot(stroke_final, aes(x=gender))+
  geom_bar(aes(fill = gender), show.legend = FALSE)+
  labs(x="",y="", title = "Sex of the patient")+
  scale_y_continuous(breaks = seq(0, 3000, by = 500), limits = c(0, 3000))+
  theme_pubclean()+
  geom_text(aes(label = paste0(after_stat(count), " (", scales::percent(after_stat(prop), accuracy = .01), ")"), group=1),
            stat = "count", vjust = -0.4, colour = "black")+
  theme(axis.title = element_text(face="bold",color="black",size=13),
        axis.text = element_text(color="black",size=12),
        plot.title = element_text(hjust = 0.5, face="bold",color="black",size=13),
        panel.grid.major.x = element_blank())

 ggplot(stroke_final, aes(x=hypertension))+
  geom_bar(aes(fill = hypertension), show.legend = FALSE)+
  labs(x="",y="", title = "Patient has hypertension")+
  scale_y_continuous(breaks = seq(0, 5000, by = 1000), limits = c(0, 5000))+
  theme_pubclean()+
  geom_text(aes(label = paste0(after_stat(count), " (", scales::percent(after_stat(prop), accuracy = .1), ")"), group=1),
            stat = "count", vjust = 1.1, colour = "black")+
  theme(axis.title = element_text(face="bold",color="black",size=13),
        axis.text = element_text(color="black",size=12),
        plot.title = element_text(hjust = 0.5, face="bold",color="black",size=13),
        panel.grid.major.x = element_blank())

 ggplot(stroke_final, aes(x=heart_disease))+
  geom_bar(aes(fill = heart_disease), show.legend = FALSE)+
  labs(x="",y="", title = "Patient has heart disease")+
  scale_y_continuous(breaks = seq(0, 5000, by = 1000), limits = c(0, 5000))+
  theme_pubclean()+
  geom_text(aes(label = paste0(after_stat(count), " (", scales::percent(after_stat(prop), accuracy = .1), ")"), group=1),
            stat = "count", vjust = 1.1, colour = "black")+
  theme(axis.title = element_text(face="bold",color="black",size=13),
        axis.text.y = element_text(color="black",size=12),
        axis.text.x = element_text(color="black",size=12, angle = 0),
        plot.title = element_text(hjust = 0.5, face="bold",color="black",size=13),
        panel.grid.major.x = element_blank())
 
 
 ggplot(stroke_final, aes(x=ever_married))+
  geom_bar(aes(fill = ever_married), show.legend = FALSE)+
  labs(x="",y="", title = "Patient is married")+
  scale_y_continuous(breaks = seq(0, 3500, by = 500), limits = c(0, 3500))+
 theme_pubclean()+
  geom_text(aes(label = paste0(after_stat(count), " (", scales::percent(after_stat(prop), accuracy = .1), ")"), group=1),
            stat = "count", vjust = 1.1, colour = "black")+
  theme(axis.title= element_text(face="bold",color="black",size=13),
        axis.text = element_text(color="black",size=12),
        plot.title = element_text(hjust = 0.5, face="bold",color="black",size=13),
        panel.grid.major.x = element_blank())


 ggplot(stroke_final, aes(x=work_type))+
  geom_bar(aes(fill = work_type), show.legend = FALSE)+
  labs(x="",y="", title = "Work type")+
  scale_y_continuous(breaks = seq(0, 3100, by = 500), limits = c(0, 3100))+
  theme_pubclean()+
  geom_text(aes(label = paste0(after_stat(count), " (", scales::percent(after_stat(prop), accuracy = .1), ")"), group=1),
            stat = "count", vjust = -0.4, colour = "black")+
  theme(axis.title = element_text(face="bold",color="black",size=13),
        axis.text.y = element_text(color="black",size=12),
        axis.text.x = element_text(color="black",size=11, angle = 0),
        plot.title = element_text(hjust = 0.5, face="bold",color="black",size=13),
        panel.grid.major.x = element_blank())


 ggplot(stroke_final, aes(x=residence_type))+
  geom_bar(aes(fill = residence_type), show.legend = FALSE)+
  labs(x="",y="", title = "Residence type")+
  scale_y_continuous(breaks = seq(0, 2800, by = 400), limits = c(0, 2800))+
  theme_pubclean()+
  geom_text(aes(label = paste0(after_stat(count), " (", scales::percent(after_stat(prop), accuracy = .1), ")"), group=1),
            stat = "count", vjust = 1.5, colour = "black")+
  theme(axis.title = element_text(face="bold",color="black",size=13),
        axis.text = element_text(color="black",size=12),
        plot.title = element_text(hjust = 0.5, face="bold",color="black",size=13),
        panel.grid.major.x = element_blank())


 ggplot(stroke_final, aes(x=smoking_status))+
  geom_bar(aes(fill = smoking_status), show.legend = FALSE)+
  labs(x="",y="", title = "Smoking status")+
  scale_y_continuous(breaks = seq(0, 2000, by = 400), limits = c(0, 2000))+
  theme_pubclean()+
  geom_text(aes(label = paste0(after_stat(count), " (", scales::percent(after_stat(prop), accuracy = .1), ")"), group=1),
            stat = "count", vjust = 1.5, colour = "black")+
  theme(axis.title = element_text(face="bold",color="black",size=13),
        axis.text.y = element_text(color="black",size=12),
        axis.text.x = element_text(color="black",size=11, angle = 0),
        plot.title = element_text(hjust = 0.5, face="bold",color="black",size=13),
        panel.grid.major.x = element_blank())



```


Majority of the patients were Female (58.6%), married (65.6%), in private employment (57.2%), had no hypertension (90.3%), had no heart disease (94.6%) and had never smoked (37.0%). There was similar proportion in rural and urban residence type.


```{r histograms, out.width="50%",fig.show="hold"}

ggplot(stroke_final, aes(x=age)) + 
  geom_histogram(color="black", fill="lightblue")+
  geom_vline(aes(xintercept=mean(age)),
            color="blue", linetype="dashed", size=1)+
   scale_y_continuous(breaks = seq(0, 300, by = 50), limits = c(0, 300))+
  scale_x_continuous(n.breaks = 10)+
  labs(x="Age of the Patient",y="count", title = "Age (Years)")+
  theme(axis.title = element_text(color="black",size=12),
        axis.text = element_text(color="black",size=11),
        plot.title = element_text(hjust = 0.5, face="bold",color="black",size=13),
        panel.grid.minor.y = element_blank(),
        panel.grid.minor.x = element_blank())

ggplot(stroke_final, aes(x=avg_glucose_level)) + 
  geom_histogram(color="black", fill="pink")+
  geom_vline(aes(xintercept=mean(avg_glucose_level)),
            color="blue", linetype="dashed", size=1)+
  scale_y_continuous(breaks = seq(0, 700, by = 100), limits = c(0, 700))+
  scale_x_continuous(n.breaks = 12)+
  labs(x="Average glucose level in blood",y="count", title = "Glucose level in blood")+
  theme(axis.title = element_text(color="black",size=12),
        axis.text = element_text(color="black",size=11),
        plot.title = element_text(hjust = 0.5, face="bold",color="black",size=13),
        panel.grid.minor.y = element_blank(),
        panel.grid.minor.x = element_blank())

ggplot(stroke_final%>%
         drop_na(bmi), aes(x=bmi)) + 
  geom_histogram(color="black", fill="lightgreen")+
  geom_vline(aes(xintercept=mean(bmi)),
            color="blue", linetype="dashed", size=1)+
    scale_y_continuous(breaks = seq(0, 900, by = 100), limits = c(0, 900))+
  scale_x_continuous(n.breaks = 10)+
  labs(x="Body mass index (in kg/m2)",y="count", title = "BMI")+
  theme(axis.title = element_text(color="black",size=12),
        axis.text = element_text(color="black",size=11),
        plot.title = element_text(hjust = 0.5, face="bold",color="black",size=13),
        panel.grid.minor.y = element_blank(),
        panel.grid.minor.x = element_blank())



```


## Bivariate analysis

### Correlation

Pearson correlation evaluates the linear relationship between two continuous variables.

```{r, fig.width=14, fig.height=10}
# improved correlation matrix
library(corrplot)

corrplot(cor(stroke_final%>%
               dplyr::select(-id)%>%
               mutate(across(c(1:11), as.numeric)),
             method='pearson', use = "complete.obs"),
  method = "color", #number
  addCoef.col = "black",
  number.cex = 0.95,
  type = "upper" # show only upper side #full
)
```


The variables **gender**, **age**, **hypertension**, **heart disease**, **ever married**, **work type**, **residence type**, **average glucose level** and **bmi** exhibit a positive linear association with our target variable (stroke) while **smoking status** a negative linear association.


```{r}
# correlation tests for whole dataset
library(Hmisc)
res <- rcorr(as.matrix(stroke_final%>%
               dplyr::select(-id)%>%
         drop_na(bmi)%>%
               mutate(across(c(1:11), as.numeric)))) # rcorr() accepts matrices only

# display p-values (rounded to 3 decimals)
kable(
  round(res$P, 3)
)
```


The linear association of stroke with **gender** and **residence type** is not significant (p>0.05). 


### Difference Frequency table


```{r}

# make dataset with variables to summarize

tbl_summary(stroke_final%>%
               dplyr::select(-id),
              by = stroke,
                       type = list(
                         all_dichotomous() ~ "categorical",
                          all_continuous() ~ "continuous2")
                       , statistic = all_continuous() ~ c(
                                      "{mean} ({sd})", 
                                      "{median} ({p25}, {p75})", 
                                      "{min}, {max}")
                       , digits = all_continuous() ~ 2
                       , missing = "always" # don't list missing data separately
                       ,missing_text = "Missing"
                       ) %>% 
   modify_header(label = "**Variables**") %>% # update the column header
   bold_labels() %>%
   italicize_levels()%>%
   add_n()%>% # add column with total number of non-missing observations
   add_p(pvalue_fun = ~style_pvalue(.x, digits = 3),
         test.args = c(work_type) ~ list(simulate.p.value=TRUE)) %>%
   bold_p(t= 0.05) %>% # bold p-values under a given threshold (default 0.05)
   #add_overall() %>%
   #add_difference() %>% #add column for difference between two group, confidence interval, and p-value
   modify_spanning_header(c("stat_1", "stat_2") ~ "**Stroke**")  %>%
   #modify_caption("**Table 1. Patient Characteristics**")%>%
   modify_footnote(
     all_stat_cols() ~ "Mean (SD); Median (IQR); Range; Frequency (%)"
   )

```


### Visualization


```{r, out.width="50%",fig.show="hold"}

library(ggthemes)

  ggplot(stroke_final, aes(x=gender))+
  geom_bar(aes(fill=stroke), position = "fill")+
  labs(title="Sex of the patient", y="Percent", x="")+
  scale_fill_brewer(palette = "Set1")+
  scale_y_continuous(breaks=seq(0,1,by=.2),label=scales::percent)+
  theme_fivethirtyeight()+
  theme(axis.title.y = element_blank(),
        #legend.title = element_blank(),
        plot.margin = unit(c(1,1,0,1),"cm"),
        axis.text = element_text(size = 11),
        plot.title = element_text(size=16,hjust = 0.5),
         panel.grid.major.x = element_blank())
  
  
  ggplot(stroke_final, aes(x=hypertension))+
  geom_bar(aes(fill=stroke), position = "fill")+
  labs(title="Patient has hypertension", y="Percent", x="")+
  scale_fill_brewer(palette = "Set2")+
  scale_y_continuous(breaks=seq(0,1,by=.2),label=scales::percent)+
  theme_fivethirtyeight()+
  theme(axis.title.y = element_blank(),
        #legend.title = element_blank(),
        plot.margin = unit(c(1,1,0,1),"cm"),
        axis.text = element_text(size = 11),
        plot.title = element_text(size=16,hjust = 0.5),
         panel.grid.major.x = element_blank())
  
  
  ggplot(stroke_final, aes(x=heart_disease))+
  geom_bar(aes(fill=stroke), position = "fill")+
  labs(title="Patient has heart disease", y="Percent", x="")+
  scale_fill_brewer(palette = "Set3")+
  scale_y_continuous(breaks=seq(0,1,by=.2),label=scales::percent)+
  theme_fivethirtyeight()+
  theme(axis.title.y = element_blank(),
        #legend.title = element_blank(),
        plot.margin = unit(c(1,1,0,1),"cm"),
        axis.text = element_text(size = 11),
        plot.title = element_text(size=16,hjust = 0.5),
         panel.grid.major.x = element_blank())
  
  
  ggplot(stroke_final, aes(x=ever_married))+
  geom_bar(aes(fill=stroke), position = "fill")+
  labs(title="Patient is married", y="Percent", x="")+
  scale_fill_brewer(palette = "Accent")+
  scale_y_continuous(breaks=seq(0,1,by=.2),label=scales::percent)+
  theme_fivethirtyeight()+
  theme(axis.title.y = element_blank(),
        #legend.title = element_blank(),
        plot.margin = unit(c(1,1,0,1),"cm"),
        axis.text = element_text(size = 11),
        plot.title = element_text(size=16,hjust = 0.5),
         panel.grid.major.x = element_blank())
  
  
  ggplot(stroke_final, aes(x=work_type))+
  geom_bar(aes(fill=stroke), position = "fill")+
  labs(title="Work type", y="Percent", x="")+
  scale_fill_brewer(palette = "Dark2")+
  scale_y_continuous(breaks=seq(0,1,by=.2),label=scales::percent)+
  theme_fivethirtyeight()+
  theme(axis.title.y = element_blank(),
        #legend.title = element_blank(),
        plot.margin = unit(c(1,1,0,1),"cm"),
        axis.text = element_text(size = 11),
        plot.title = element_text(size=16,hjust = 0.5),
         panel.grid.major.x = element_blank())
  
  
  ggplot(stroke_final, aes(x=residence_type))+
  geom_bar(aes(fill=stroke), position = "fill")+
  labs(title="Residence type", y="Percent", x="")+
  scale_fill_brewer(palette = "Paired")+
  scale_y_continuous(breaks=seq(0,1,by=.2),label=scales::percent)+
  theme_fivethirtyeight()+
  theme(axis.title.y = element_blank(),
        #legend.title = element_blank(),
        plot.margin = unit(c(1,1,0,1),"cm"),
        axis.text = element_text(size = 11),
        plot.title = element_text(size=16,hjust = 0.5),
         panel.grid.major.x = element_blank())
  
  
  ggplot(stroke_final, aes(x=smoking_status))+
  geom_bar(aes(fill=stroke), position = "fill")+
  labs(title="Smoking status", y="Percent", x="")+
  scale_fill_brewer(palette = "Spectral")+
  scale_y_continuous(breaks=seq(0,1,by=.2),label=scales::percent)+
  theme_fivethirtyeight()+
  theme(axis.title.y = element_blank(),
        #legend.title = element_blank(),
        plot.margin = unit(c(1,1,0,1),"cm"),
        axis.text = element_text(size = 11),
        plot.title = element_text(size=16,hjust = 0.5),
         panel.grid.major.x = element_blank())

```


From the graphs, there is no difference in proportion of stroke with **gender** and **residence type**.


```{r, fig.width=14, fig.height=10}

p1 <- ggplot(stroke_final, aes(x=age, fill=stroke)) + 
  geom_density(alpha=0.4)+
  scale_y_continuous(n.breaks = 10)+
  scale_x_continuous(breaks = seq(0, 85, by = 5), limits = c(0, 85))+
  labs(x="age",y="density", title = "Age of the patient")+
  theme(axis.title = element_text(color="black",size=12),
        legend.position = "top",
        axis.text = element_text(color="black",size=11),
        plot.title = element_text(hjust = 0.5, face="bold",color="black",size=13),
        panel.grid.minor.y = element_blank(),
        panel.grid.minor.x = element_blank())+
  guides(fill = guide_legend(title = "Stroke"))


p2 <- ggplot(stroke_final, aes(x=avg_glucose_level, fill=stroke)) + 
  geom_density(alpha=0.4)+
  scale_y_continuous(n.breaks = 10)+
  scale_x_continuous(breaks = seq(50, 300, by = 25), limits = c(50, 300))+
  labs(x="Average glucose level in blood",y="density", title = "Glucose level in blood")+
  theme(axis.title = element_text(color="black",size=12),
        legend.position = "top",
        axis.text = element_text(color="black",size=11),
        plot.title = element_text(hjust = 0.5, face="bold",color="black",size=13),
        panel.grid.minor.y = element_blank(),
        panel.grid.minor.x = element_blank())+
  guides(fill = guide_legend(title = "Stroke"))


p3 <- ggplot(stroke_final, aes(x=bmi, fill=stroke)) + 
  geom_density(alpha=0.4)+
  scale_y_continuous(n.breaks = 10)+
  scale_x_continuous(breaks = seq(0, 100, by = 10), limits = c(0, 100))+
  labs(x="Body mass index (in kg/m2)",y="density", title = "BMI")+
  theme(axis.title = element_text(color="black",size=12),
        legend.position = "top",
        axis.text = element_text(color="black",size=11),
        plot.title = element_text(hjust = 0.5, face="bold",color="black",size=13),
        panel.grid.minor.y = element_blank(),
        panel.grid.minor.x = element_blank())+
  guides(fill = guide_legend(title = "Stroke"))


p4 <- ggplot(stroke_final, aes(stroke, age))+
  geom_boxplot(aes(colour = stroke), outlier.colour = "black", 
               outlier.shape = 8, show.legend = TRUE)+
  labs(x="",y="age", title = "Age of the patient")+
  scale_y_continuous(breaks = seq(0, 90, by = 10), limits = c(0, 90))+
  theme(axis.title = element_text(face="bold",color="black",size=13),
        legend.position = "bottom",
        axis.text = element_text(color="black",size=12),
        plot.title = element_text(hjust = 0.5, face="bold",color="black",size=13),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.y = element_blank())


p5 <- ggplot(stroke_final, aes(stroke, avg_glucose_level))+
  geom_boxplot(aes(colour = stroke), outlier.colour = "black", 
               outlier.shape = 1, show.legend = TRUE)+
  labs(x="",y="Average glucose level in blood", title = "Glucose level in blood")+
  scale_y_continuous(breaks = seq(0, 300, by = 25), limits = c(0, 300))+
  theme(axis.title = element_text(face="bold",color="black",size=13),
        legend.position = "bottom",
        axis.text = element_text(color="black",size=12),
        plot.title = element_text(hjust = 0.5, face="bold",color="black",size=13),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.y = element_blank())


p6 <- ggplot(stroke_final, aes(stroke, bmi))+
  geom_boxplot(aes(colour = stroke), outlier.colour = "black", 
               outlier.shape = 1, show.legend = TRUE)+
  labs(x="",y="Body mass index (in kg/m2)", title = "BMI")+
  scale_y_continuous(breaks = seq(0, 100, by = 10), limits = c(0, 100))+
  theme(axis.title = element_text(face="bold",color="black",size=13),
        legend.position = "bottom",
        axis.text = element_text(color="black",size=12),
        plot.title = element_text(hjust = 0.5, face="bold",color="black",size=13),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.y = element_blank())


figure1 <- ggarrange( p1, p2, p3, p4, p5, p6, 
          ncol = 3, nrow = 2)

annotate_figure(figure1,
                top = text_grob("",
                                color = "red", face = "bold", size = 15),
                bottom = text_grob("", color = "blue",
                                   hjust = 1, x = 0.98, face = "italic", size = 10),
                #left = text_grob("Figure arranged using ggpubr", color = "green", rot = 90),
                #right = "",
                fig.lab = "", fig.lab.face = "bold"
                )

```


# Modelling

From the exploration done, it is very hard to make a conclusion. The dependent variable “stroke” is imbalanced. Patient had stroke (4.9%) vs Patient who did no have stroke (95.1%). It will be very wrong to just perform a predictive model on such kind of data without making any changes.

Before creating any model, the first step is to drop missing cases then divide the data into training and testing dataset.

```{r}
stroke_model <- stroke_final%>%
  drop_na()

table(stroke_model$stroke)

```


```{r}

library(caTools)

set.seed(123) 

sample <- sample.split(stroke_model$stroke, SplitRatio = 0.8)


train <- subset(stroke_model, sample==TRUE)


test <- subset(stroke_model, sample==FALSE)


#train=stroke_model[sample==TRUE, ]

#test=stroke_model


table(train$stroke)

table(test$stroke)

```


## Random Forest - categorical

### Imbalanced Data

```{r}

library(randomForest)

model1_rf <- randomForest(stroke ~., data = train)
model1_rf



```

```{r}


library(caret)
confusionMatrix(predict(model1_rf, test), test$stroke, positive = "Yes")

```

From the above output we can see that the accuracy of the model is 95.72%. 

However we need to be sure that this model is good by checking the sensitivity and the specificity. Sensitivity is the accuracy of predicting what we are interested in, that is a person having experienced a stroke while specificity is now for no stroke. 

From the output the sensitivity is at 0% while specificity is 100%. This means that this model will be more accurate in predicting those who did not have stroke than those who did. This problem has been caused by the imbalance of the dependent variable. 

This problem can be solved in several ways including; over sampling, undersampling and both. We will use the package called ROSE.

```{r}
library(ROSE)
```


### Oversampling data

We specify N=7518 (3759x2). Over sampling repeats some values in the category with the fewer entries randomly until they are the same with the larger entry. 

```{r}

over_sampling <-ovun.sample(stroke ~.,data=train, method = "over", N=7518, seed=1)$data

table(over_sampling$stroke)

```


```{r}


model2_rf <- randomForest(stroke ~., data = over_sampling)
model2_rf


```

```{r}

confusionMatrix(predict(model2_rf, test), test$stroke, positive = "Yes")

```

From the output above, we can see that sensitivity is still 0% while specificity has gone down to 99.57%. This proportion is not still good.


### Under sampling data

This reduces the entries in the category with more entries to be the same with the one with few entries hence (167*2=334).

```{r}

under_sampling <-ovun.sample(stroke ~.,data=train, method = "under", N=334, seed=2)$data

table(under_sampling$stroke)

```

```{r}


model3_rf <- randomForest(stroke ~., data = under_sampling)
model3_rf


```



```{r}

confusionMatrix(predict(model3_rf, test), test$stroke, positive = "Yes")

```

From the output above, sensitivity now stands at 76.19% while specificity stands at 72.13%. This is a better model than the unbalanced and over sampled model.

### Both Sampling data

```{r}

both_sampling <- ovun.sample(stroke~.,data=train, method="both",p=0.5, seed=222, N=3926)$data

table(both_sampling$stroke)

```

```{r}


model4_rf <- randomForest(stroke ~., data = both_sampling)
model4_rf


```

```{r}

confusionMatrix(predict(model4_rf, test), test$stroke, positive = "Yes")

```

From the output above, the sensitivity now is at 23.81% while specificity is at 97.23%.

- This means that the under sampled model is the best to predict stroke with accuracy of 73.52%, sensitivity of 78.57% and specificity of 73.29%.  


## Tree





